Three $\approx$40 Million Year Old Mini-Neptunes from Kepler, TESS, and Gaia

### 2022-03-28 papers/ms.tex
update ms

%
% Symbols
%
\newcommand{\kms}{\,km\,s$^{-1}$}
\newcommand{\ms}{\,m\,s$^{-1}$}
\newcommand{\bpmrpo}{(G_{\rm BP}-G_{\rm RP})_0}
\newcommand{\bpmrp}{G_{\rm BP}-G_{\rm RP}}

@@

In this work, we leverage recent analyses of the Gaia data, which have
greatly expanded our knowledge of stellar group memberships
\citep[{\it e.g.},][]{CantatGaudin2018a,KounkelCovey2019,Kerr2021}.
Broadly speaking these analyses cluster on the stellar positions and
on-sky velocities measured by Gaia, with varying degrees of filtering
and supervision.  One important result is the identification of
diffuse streams and tidal tails comparable in stellar mass to the
previously known cores of nearby open clusters
\citep{meingast_psceri_2019,Meingast2021,gagne_number_2021}.  These
spatially diffuse groups typically have velocity dispersions of
$\approx$1\kms, though they can be much higher due to both projection
effects and internal dynamics.  As an extreme example, in the Hyades
the apparent velocities of stars in the tidal tails are expected to
span up to $\pm 40$\kms relative to the cluster center
\citep{jerabkova_800_2021}.  The stars in such diffuse regions can be
verified to be the same age as the core members ({\it i.e.}, co-eval)
through analyses of color-absolute magnitude diagrams, stellar
rotation periods \citep{curtis_tess_2019,bouma_2021_ngc2516}, and
chemical abundances \citep{hawkins_2020}.  While there are many
implications for our understanding of star formation and cluster
evolution \citep[{\it e.g.}][]{dinnbier_tidal_2020}, a more immediate
consequence is that we now know the ages of many more stars, including
previously known planet hosts.

@@

The CAMDs show that for RSG-5, our membership selection gives stars
that photometrically appear to almost all be on a tight
pre-main-sequence locus.  This implies a false positive rate of at
most a few percent.  By comparison, our control sample (the $\delta$~Lyr
candidates) has a false positive rate of $\approx$12\%, based on the
number of stars that photometrically appear to be more consistent with
the field population than the bulk cluster population.  For CH-2, our
membership selection gives 27 objects in the color range displayed,
and 23 of them appear to be consistent with being on the
pre-main-sequence.  This implies a false positive rate of
$\approx$15\%.

@@

An independent way to assess the age of the candidate cluster members
is to measure their stellar rotation periods using data available from
surveys such as TESS \citep{ricker_transiting_2015} and ZTF
\citep{bellm_zwicky_2019}.  This approach leverages a storied
tradition of measuring rotation periods for open clusters of various
ages, in order to determine the rate of stellar spin-down \citep[{\it
e.g.},][]{skumanich_time_1972,curtis_rup147_2020}.  The TESS data in
our case are especially useful, since they provide 3 to 5 lunar months
of photometry for all of our candidate CH-2 and RSG-5 members.

We selected stars suitable for gyrochronology by requiring $\bpmrpo
\geq 0.5$ and $G<16$.  The latter cut corresponds to $\bpmrpo \lesssim
2.6$, at the relevant distances.  These cuts gave 19 stars in CH-2 and
42 stars in RSG-5.  We extracted light curves from the TESS images for
these stars using the \texttt{unpopular} package
\citep{hattorio_2021_cpm}, and regressed them against systematics with
its causal pixel model.  
{\bf For the ZTF light curves, we used the default photometry?}.
We then measured candidate rotation periods
using a Lomb-Scargle periodogram, and visually inspected them
following the methods discussed in \citet{curtis_rup147_2020}.

The lower panels of Figure~\ref{fig:age} show the results.
In RSG-5,
36/42 stars have rotation consistent with ``sub-Pleiades'' (86\%).
Given the nearly perfectly clean photometric CAMD for RSG-5, this
probably gives some indication for the fraction of stars for which we
should expect
to detect rotation periods with TESS (due to e.g., pole-on stars
having less detectable starspot modulation).

Consider stars in CH-2 with 0.5 < BP-RP0, and G < 16.
19 stars meet this criterion -- all were observed by TESS.
13/19 (68\%) have rotation that is obviously consistent with a 40 Myr
age.
4 are M-dwarfs with rotation periods between 10 and 12.5 days, for which
the age interpretation is not obvious.
L.~Rebull for instance have found numerous M-dwarfs
with 10-12 day rotation periods at ages of USco ($\sim 8$\,Myr); some
might exist at ages of LCC ($\sim 16$\,Myr)
(\citealt{rebull_rotation_2020}; Rebull et al., in prep -- {\bf see
Skumanich conf slide deck -- though beware TESS stitching systematics}).
Fr the sake of argument, let's say that this provides an independent
verification that at least two thirds of our CH-2 candidates are both
photometrically and rotationally consistent with being $\approx$40
Myr.

For RSG-5, 42 stars are in the relevant color and brgithness range.
36 have rotation consistent with ``sub-Pleiades'', i.e., 86\%.
Given the nearly perfectly clean photometric CAMD for RSG-5, this
probably gives some indication for the fraction of stars for which we
should expect
to detect rotation periods with TESS (due to e.g., pole-on stars
having less detectable starspot modulation).

@@

>>>>

%
% Symbols
%
\newcommand{\kms}{\,km\,s$^{-1}$}
\newcommand{\mkms}{\,km\,s^{-1}}  % math mode
\newcommand{\ms}{\,m\,s$^{-1}$}
\newcommand{\bpmrpo}{(G_{\rm BP}-G_{\rm RP})_0}
\newcommand{\bpmrp}{G_{\rm BP}-G_{\rm RP}}

@@

In this work, we leverage recent analyses of the Gaia data, which have
greatly expanded our knowledge of stellar group memberships
\citep[{\it e.g.},][]{CantatGaudin2018a,KounkelCovey2019,Kerr2021}.
Broadly speaking these analyses cluster on the stellar positions and
on-sky velocities measured by Gaia, with varying degrees of filtering
and supervision.  One important result is the identification of
diffuse streams and tidal tails comparable in stellar mass to the
previously known cores of nearby open clusters
\citep{meingast_psceri_2019,Meingast2021,gagne_number_2021}.  These
spatially diffuse groups typically have velocity dispersions of
$\approx$1\kms, though they can be much higher due to both projection
effects and internal dynamics.  As an extreme example, in the Hyades
the apparent velocities of stars in the tidal tails are expected to
span up to $\pm 40$\kms relative to the cluster center
\citep{jerabkova_800_2021}.  The stars in such diffuse regions can be
verified to be the same age as the core members ({\it i.e.}, coeval)
through analyses of color-absolute magnitude diagrams, stellar
rotation periods \citep{curtis_tess_2019,bouma_2021_ngc2516}, and
chemical abundances \citep{hawkins_2020}.  While there are many
implications for our understanding of star formation and cluster
evolution \citep[{\it e.g.},][]{dinnbier_tidal_2020}, a more immediate
consequence is that we now know the ages of many more stars, including
previously known planet hosts.

@@

The CAMDs show that for RSG-5, our membership selection gives stars
that photometrically seem to almost all be on a tight
pre-main-sequence locus.  This implies a false positive rate of 
a few percent, at most.  By comparison, our control sample (the $\delta$~Lyr
candidates) has a false positive rate of $\approx$12\%, based on the
number of stars that photometrically appear to be more consistent with
the field population than the bulk cluster population.  For CH-2, our
membership selection gives 27 objects in the color range displayed,
and 23 of them appear to be consistent with being on the
pre-main-sequence.  This implies a false positive rate of
$\approx$15\%.

@@

An independent way to assess the age of the candidate cluster members
is to measure their stellar rotation periods.  This approach can be
achieved using surveys such as TESS \citep{ricker_transiting_2015} and
ZTF \citep{bellm_zwicky_2019}, and it leverages a storied tradition of
rotation period measurement for benchmark open clusters \citep[see
{\it e.g.},][]{skumanich_time_1972,curtis_rup147_2020}.  The TESS data
in our case are especially useful, since they provide 3 to 5 lunar
months of photometry for all of our candidate CH-2 and RSG-5 members.

We selected stars suitable for gyrochronology by requiring $\bpmrpo
\geq 0.5$ and $G<16$.  The latter cut corresponds to $\bpmrpo \lesssim
2.6$, at the relevant distances.  These cuts gave 19 stars in CH-2 and
42 stars in RSG-5.  We extracted light curves from the TESS images for
these stars using the \texttt{unpopular} package
\citep{hattorio_2021_cpm}, and regressed them against systematics with
its causal pixel model.  {\bf For the ZTF light curves, we used the
default photometry?  Ran aperture photometry on the image cutouts?}.
We then measured candidate rotation periods using a Lomb-Scargle
periodogram, and visually inspected them following the methods
discussed in \citet{curtis_rup147_2020}.

The lower panels of Figure~\ref{fig:age} show the results.  In RSG-5,
36/42 stars have rotation periods faster than the Pleiades (86\%).
This numerator omits the two stars with periods $>12$\,days visible in
the lower-left panel of Figure~\ref{fig:age}.  The age interpretation
for these latter stars, particularly the $\approx$M2.5 dwarf, is not
obvious.  \citet{rebull_usco_2018} for instance have found numerous
M-dwarfs with 10-12 day rotation periods at ages of USco ($\sim
8$\,Myr), and some may still exist at ages of LCC ($\sim 16$\,Myr;
L.~Rebull in preparation).  Regardless, given that nearly no field
star outliers seem to be present on the RSG-5 CAMD, the fact that we
do not detect rotation periods for $\approx$14\% of stars should
perhaps be taken as an indication for the fraction of stars for
rotation periods might not be detectable, due to {\it e.g.}, pole-on
stars having lower amplitude starspot modulation.

%
% CH-2
% 5 stars with Prot>10 days (from the Prot versus color plot):
%                      DESIGNATION  TESS_Data   bp_rp_x     period
%25  Gaia EDR3 2129930258400157440          4  1.909693  15.309000
%24  Gaia EDR3 2129417404945830784          4  2.040988  10.981582
%4   Gaia EDR3 2107169165113864064          4  2.059904  10.580494
%9   Gaia EDR3 2134851775526125696          4  2.326998  12.227900
%1   Gaia EDR3 2127726081184588160          4  2.347244  11.543281
%
% CH-2
% Photometric field stars (from Glue -- one of these is omitted due to phot S/N cut, 
% presumably 2126008609657702784 )
%
%In [9]: df[['DESIGNATION', 'bp_rp', 'M_G']]
%Out[9]:
%                     DESIGNATION     bp_rp        M_G
%0  Gaia EDR3 2126529468937923456  2.518077   9.844733 --> NaN period, G=16.9
%1  Gaia EDR3 2126008609657702784  2.746735  11.409737 --> NaN, G=18.4
%2  Gaia EDR3 2133330051428541952  2.499458   9.459546 --> 15.7 d (ZTF), G=16.5
%3  Gaia EDR3 2133510921093964928  2.720177  10.923209 --> NaN, G=17.9
%4  Gaia EDR3 2139369183468814464  2.929506  11.170221 --> NaN, G=18.2
%5  Gaia EDR3 2129930258400157440  1.909693   8.200652 --> P=15.3 (ZTF), G=15.2

For CH-2, 13/19 stars have rotation periods that are obviously faster
than their counterparts in the Pleiades (68\%).  4 stars, not included
in the preceding numerator, are M-dwarfs with rotation periods between
10 and 12.5 days.  The age interpretation for these M-dwarfs is, as
just discussed, not obvious.  Regardless, the $\approx$15\% false
positive rate determined from the CAMD seems consistent with our
fraction of detected rotation periods, given that RSG-5 was also
missing rotation period detections for $\approx$15\% of its candidate
members, which all seemed photometrically consistent with being part
of a single pre-main-sequence locus.

It is challenging to convert these stellar rotation periods
to a precise age estimate, since on the pre-main-sequence
the stars are spinning up due to thermal contraction
rather than down due to magnetized braking.  Regardless, the rotation
period distributions of both CH-2 and RSG-5 seem consistent with other
30\,Myr to 50\,Myr clusters ({\it e.g.}, IC\,2602 and IC\,2391;
\citealt{douglas_stephanie_t_2021_5131306}).
They also seem consistent with the false positive rate estimates
determined from the color-absolute magnitude diagrams.

@@

\subsection{Is CH-2 really a star cluster?}

RSG-5, and Kepler-1643's membership inside it, clearly meet the
typical expectations of a star claimed to be in an open cluster.
RSG-5 shows an obvious overdensity relative to the local field
population ({\it e.g.}, Figure~\ref{fig:XYZvtang}), and our membership
selection easily produces a clean pre-main-sequence locus in 
color-absolute magnitude space (Figure~\ref{fig:age}).
CH-2, and KOI-7913 and KOI-7368's membership inside it, do not meet
those expectations in as obvious a manner.  This is because this
association of stars is diffuse.

%really 80:9
To quantify the density discrepancy, we can compare the spatial and
velocity volumes searched to select candidate members of each cluster.
For RSG-5, we drew 141 candidate members from a $30\,{\rm
pc}\times30\,{\rm pc}\times30\,{\rm pc}$ spatial cube, given a $1 {\rm \mkms} \times 2 {\rm \mkms }$ rectangle in apparent galactic velocity.
For CH-2, our 37 candidate members came from a spatial cube of
dimension $50\,{\rm pc}\times40\,{\rm pc}\times30\,{\rm pc}$, and the
velocity rectangle of $2 {\rm \mkms} \times 4 {\rm \mkms}$.  If we
define the ``searched volume'' in units of ${\rm pc}^3 {\rm
(\mkms)}^2$, then the volume ratio of CH-2:RSG-5 is $\approx$9:1.  The
density (number of stars per unit searched volume) within RSG-5 relative to
CH-2 similarly comes out to 34 to 1.

So what makes a star cluster?  Historic answers to this question have
recently by reviewed by \citet{krumholz_star_2019}: some definitions
that have been offered include criteria such as being gravitationally
bound, and having a mass density that significantly exceeds the mean
in a cluster's galactic neighborhood.  We prefer a modified version of
the definition adopted by \citet{krumholz_star_2019}: for our purposes
a star cluster is a group of at least 12 stars that was physically
associated at its time of formation.  The somewhat arbitrary ``12'' is
set to distinguish clusters from high-order multiple star systems.  We
therefore explicitly include dissolved clusters and their tidal tails
in our concept of clusters.  We also explicitly exclude the idea that
a particular number of stars per unit spatial volume is required to
define a cluster.  The latter point acknowledges the fact that an
important factor in cluster identification is now also the number per
unit velocity volume, whether in 2-dimensional tangential velocity, or
when including the third radial component.  Perhaps
once stellar rotation periods and chemical abundances reach the same
level of ubiquity as stellar proper motions, they might enable further
refinement of our ability in cluster discovery.

From a data-driven perspective, how do we demonstrate that a star is
in a cluster, {\it i.e.}, that it is part of a group of stars that was
physically associated at its time of formation?
Back-integrating the orbits is one convincing approach, but it does
not always work {\bf (CITE)}.
The relatively minimal approach suggested by \citet{tofflemire_2021}
is intriguing: search for coeval, phase-space neighbors, measure their
ages, and determine if they share a common age.
This approach is more accurately be described as a method for
determining whether a star is currently associated with a set of coeval
stars, which is much easier to determine than what the association
looked like in the past.
From this standard, our analysis thus far of CH-2 has already
demonstrated the existence of such an assocation.

A crucial logical step in this method however is to ensure that the
(automated) search process for coeval phase-space neighbors in fact
produces neighbors at a rate different from what it would for field
stars.

% TODO
And so we ask: what is the density of field stars in the CH-2 region?
If we had applied Ronan's pre-main-sequence selection, + HDBScan, +
local neighbors... what would we have gotten?
Similarly... if you just applied Adam Kraus' Comove 
What is the likelihood we are fooling ourselves?
